%% ===================================================================
\section{Experimental Results}
\label{sec:eval}

\subsection{PolyBench/C (30 Kernels)}

\begin{table}[t]
  \caption{PolyBench/C results.  WA = with analysis, NA = no analysis.}
  \label{tab:results_1x}
  \centering\small
  \begin{tabular}{lcc}
    \toprule
    \textbf{Metric} & \textbf{WA} & \textbf{NA} \\
    \midrule
    Pass rate              & 29/30 (97\%) & 28/30 (93\%) \\
    First-try passes       & 12            & 8             \\
    Median speedup         & 0.76$\times$  & 0.88$\times$ \\
    Mean speedup           & 1.24$\times$  & 1.64$\times$ \\
    Kernels $>1\times$     & 11/29         & 12/28         \\
    \midrule
    \multicolumn{3}{l}{\emph{Head-to-head (28 common):}} \\
    WA wins ($>$10\% faster) & \multicolumn{2}{c}{12} \\
    NA wins ($>$10\% faster) & \multicolumn{2}{c}{9}  \\
    Ties (within 10\%)       & \multicolumn{2}{c}{8}  \\
    WA-only pass             & \multicolumn{2}{c}{1 (jacobi\_2d)}  \\
    \bottomrule
  \end{tabular}
\end{table}

Table~\ref{tab:results_1x} summarizes results at standard
(1$\times$) problem sizes.  With analysis guidance, 29/30 kernels pass
correctness (1 failure due to LLM nondeterminism); without it, 28/30 pass but analysis uniquely enables \texttt{jacobi\_2d}.
The median speedup at 1$\times$ sizes is limited by PolyBench's
small defaults ($N$\,=\,60--120), where GPU launch overhead dominates.
Compute-bound kernels show strong speedups
(\texttt{3mm}: 4.01$\times$, \texttt{covariance}: 3.66$\times$,
\texttt{deriche}: 3.58$\times$, \texttt{2mm}: 2.97$\times$).

In the head-to-head comparison, WA wins 12 of 28 common kernels while NA wins 9. The NA wins are largely due to LLM nondeterminism (different code quality across runs) rather than a systematic advantage of unguided generation.

\paragraph{Scaling.}
At 8$\times$ sizes ($N$: 60$\to$480, matrices: 120$\to$960),
27/30 pass with median speedup of 24.57$\times$ over single-threaded C,
confirming that GPU advantage grows with problem size.
Top kernels: \texttt{3mm} (2041$\times$), \texttt{2mm} (2004$\times$),
\texttt{gemm} (451$\times$), \texttt{syrk} (62$\times$).

\subsection{Profiling-Guided Optimization}

\begin{table}[t]
  \caption{Representative profiling feedback improvements (PolyBench).
    ``Before'' = first correct kernel; ``After'' = best after NCU-guided iterations.}
  \label{tab:profiling}
  \centering\small
  \begin{tabular}{lccl}
    \toprule
    \textbf{Kernel} & \textbf{Before} & \textbf{After} & \textbf{Scale} \\
    \midrule
    \multicolumn{4}{l}{\emph{1$\times$ sizes:}} \\
    3mm         & 0.10$\times$ & 4.01$\times$ & 1$\times$ \\
    covariance  & 0.11$\times$ & 3.66$\times$ & 1$\times$ \\
    2mm         & 1.48$\times$ & 2.97$\times$ & 1$\times$ \\
    syrk        & 1.65$\times$ & 2.07$\times$ & 1$\times$ \\
    \midrule
    \multicolumn{4}{l}{\emph{8$\times$ sizes:}} \\
    gemm        & 120$\times$  & 451$\times$  & 8$\times$ \\
    correlation & 12$\times$   & 51$\times$   & 8$\times$ \\
    atax        & 0.4$\times$  & 15$\times$   & 8$\times$ \\
    syrk        & 24$\times$   & 62$\times$   & 8$\times$ \\
    \bottomrule
  \end{tabular}
\end{table}

Profiling feedback improved 9 of 29 kernels at 1$\times$ and 12 of 21 benchmarked kernels at 8$\times$ (Table~\ref{tab:profiling}).
The gains are largest at 8$\times$ sizes where hardware utilization bottlenecks are more pronounced: \texttt{gemm} improves from 120$\times$ to 451$\times$ as the LLM restructures for better memory coalescing after receiving the memory-bound diagnosis.
Correctness is preserved: each optimized kernel is re-validated, and the majority of optimization attempts that break correctness are automatically rejected.

\subsection{TSVC (151 Kernels)}

\sysname passes 143/151 TSVC kernels (94.7\%), 91 on first attempt.
At 32\,KB--32\,MB per array, LLM-generated Triton achieves a mean speedup of 2.78$\times$ over single-threaded C,
with a maximum of 267$\times$.

\subsection{Application Kernels}

\begin{table}[t]
  \caption{Application kernels: Triton speedup vs.\ single-threaded C.}
  \label{tab:realworld}
  \centering\small
  \begin{tabular}{llcc}
    \toprule
    \textbf{Source} & \textbf{Kernel} & \textbf{WA/ST} & \textbf{Pass?} \\
    \midrule
    ECP     & LJ force   & 24.1$\times$ & Y \\
    ECP     & SpMV       & 7.1$\times$  & Y \\
    Rodinia & SRAD       & 12.8$\times$ & Y \\
    Rodinia & pathfinder & 3.6$\times$  & Y \\
    Rodinia & gaussian   & 0.6$\times$  & Y \\
    Rodinia & lud        & 1.3$\times$  & Y \\
    \midrule
    Rodinia & hotspot    & ---          & N \\
    Rodinia & lavaMD     & ---          & N \\
    \bottomrule
  \end{tabular}
\end{table}

We apply \sysname to 3 Rodinia and 5 ECP/Rodinia application kernels (Table~\ref{tab:realworld}).
6 of 8 pass correctness with analysis guidance.
The highest speedups come from compute-intensive kernels:
\texttt{LJ\_force} (24.1$\times$) and \texttt{SRAD} (12.8$\times$).
Two kernels fail: \texttt{hotspot} (complex 2D stencil with boundary conditions) and \texttt{lavaMD} (irregular particle interactions), both requiring domain-specific optimizations beyond what the unified analysis provides.
